To be FAIR: Theory Specification Needs an Update

### 2025-02-20 manuscript.Rmd
update 2025-02-20 12:48:57.507521
closes #37

abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
keywords          : "fairtheory, meta theory, theory formation, cumulative science, formal models"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "theory-specification.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            :
  papaja::apa6_pdf: default
  md_document: default
---

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories; others have coined *abduction* to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. However, Peirce also acknowledged that abduction "is not essentially different from induction" (p.4881). Defining what makes a theory "best" requires making auxiliary assumptions. For present purposes, induction and abduction are interchangeable terms.].
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect, and it is in turn [indexed by Google Scholar](https://support.zenodo.org/help/en-gb/29-indexing/61-is-zenodo-indexed-by-google-scholar), Google Dataset Search, and other services.
Thus, while it is fine (and recommended) to use GitHub for version control,
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

@@

From a pragmatic perspective,
these considerations underscore the importance of being as explicit as possible about assumptions and ontological definitions when making theory Accessible,
while acknowledging that Accessibility exists on a spectrum, 
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, while it is impossible to communicate a theory completely unambiguously,
scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
A theory's Accessibility is increased by
reducing dependencies on (implicit) background knowledge,
explication of assumptions,
formalization,
and explicit cross-references to relevant resources (papers, ontologies, macro-theories, measurement instruments, experimental designs) [@langeChecklist2025].

>>>>

abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
keywords          : "fairtheory, meta theory, theory formation, cumulative science, formal models"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "fair_theory.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            :
  papaja::apa6_pdf: default
  md_document: default
---

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories. Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation". However, Peirce later acknowledged that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881]. As defining what makes a theory "best" requires auxiliary assumptions, we consider abduction to be a special case of induction. For present purposes, however, the terms are interchangeable.].
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Thus, while it is fine (and recommended) to use GitHub for version control,
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

@@

From a pragmatic perspective,
these considerations underscore the importance of being as explicit as possible about assumptions and ontological definitions when making theory Accessible,
while acknowledging that Accessibility exists on a spectrum, 
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, while it is impossible to communicate a theory completely unambiguously,
scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
A theory's Accessibility is increased by
reducing dependencies on (implicit) background knowledge,
explication of assumptions,
formalization,
and explicit cross-references to relevant resources (papers, ontologies, macro-theories, measurement instruments, experimental designs) [REF langeChecklist2025].
